---
title: "Final paper results Sept 2"
author: "Kristin Dobbin"
date: "2025-09-03"
output:
  pdf_document: default
  html_document: default
knit: (function(rmdfile, ...) { rmarkdown::render(rmdfile, output_dir=here::here("Docs"))})
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = "/Users/KristinDobbin/Documents/Projects/R_Projects_Active/CAcarceralwatersystems", warning=FALSE, message=FALSE, include=FALSE)

library(tidyverse)
library(gtsummary)
library(kableExtra)
```


```{r get and prep data, include=FALSE, echo=FALSE}
Data <- read_csv("Data/Fullpaperdata_Sept.csv")
Data <- Data %>% mutate(across(c(pwsid, pws_type_code, carceral_sys, health_5yr_no2ndary_or_lsl_binary, mon_only_5yr_binary, mr_5yr_binary, CASDWIS_County, watersource, purchased, Failing_Status, SAFER_Status, Absence_of_Interties, Source_Capacity_Violations, Bottled_Water_or_Hauled_Water_Reliance, Significant_Deficiencies, SC5e_Drought_Impact, Distributiondis_any, HR_NAME), as.factor))
Data$Missing_ccr_any <- as.factor(ifelse(Data$n_ccr_missing > 0, "Yes", "No"))
Data$Missing_ccr_twoplus <- as.factor(ifelse(Data$n_ccr_missing >= 2, "Yes", "No"))
```

## Data overview

Our data set includes `r nrow(Data)` unique public water systems representing all of the active CWS and NTNCWS water systems in the same. Of these water systems `r sum(Data$pws_type_code == "CWS")` are community water systems, `r sum(Data$pws_type_code == "NTNCWS")` are Non-Transient Non-Community systems. The mean population served by these systems is `r mean(Data$population_served_count)`. 

## Describing carceral water systems

We classify these systems into three categories: Carceral water systems that primarily or exclusively serve carceral facilities;  non-exclusive carceral water systems that serve one or more carceral facilities in addition to other customers and non-carceral systems that do not serve any carceral faciltiies. We find that `r sum(Data$carceral_sys == "carceral_water_system")` or `r ((sum(Data$carceral_sys == "carceral_water_system"))/nrow(Data)*100)`% are carceral water systems.  `r sum(Data$carceral_sys == "carceral_serving_water_system")` or `r ((sum(Data$carceral_sys == "carceral_serving_water_system"))/nrow(Data)*100)`% are non-exclusive carceral systems and `r sum(Data$carceral_sys == "other_water_systems")` or `r ((sum(Data$carceral_sys == "other_water_systems"))/nrow(Data)*100)`% are noncarceral systems.

```{r describing carceral water system characteristics, echo=FALSE, warning=FALSE, message=FALSE}
System_descriptives <- Data[, c(3,6,8,17,18,19)]

Table <- System_descriptives %>% tbl_summary(by = carceral_sys)

kable_Table <- Table %>% as_kable_extra(); kable_Table
```

As you can see above, many of the carceral serving systems rely on surface water. Which is interesting. Many of those systems are concentrated in the bay area, Sacramento and northern central valley areas and southern california whereas exclusive carceral systems are fairly evenly spread throughout the state.

```{r describing carceral water system map, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(ggmap)
library(sf)
library(sp)

#load in boundary polygons for CWS
PWS_boundary <- st_read("Data/California_Drinking_Water_System_Area_Boundaries.geojson") #SABL download from Jenny 072325
str(PWS_boundary)
PWS_boundary <- st_make_valid(PWS_boundary)

PWS_boundary$SABL_PWSID <- as.factor(PWS_boundary$SABL_PWSID)
PWS_boundary$WATER_SYSTEM_NAME <- as.factor(PWS_boundary$WATER_SYSTEM_NAME)
PWS_boundary <- PWS_boundary %>% filter(BOUNDARY_TYPE != "Jurisdictional") #remove jurisdictional boundaries to get rid of duplicates

Geodata <- Data
Geodata <- left_join(Geodata, PWS_boundary, by = c("pwsid" = "SABL_PWSID"))
Geodata <- Geodata %>% distinct(pwsid, .keep_all = TRUE) #get rid of one duplicate

#get centroids
Geodata$Centroid <- NA
Geodata$Centroid <- st_centroid(Geodata$geometry) #Get centroids to service area boundaries

#Basemaps
caCountiesTmp <- tigris::counties(state = 06) %>%
  st_as_sf()

Geodata <- st_as_sf(Geodata, sf_column_name = "Centroid")

#maps - statewide
Map <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica"))

```

```{r system map, echo=FALSE, warning=FALSE, message=FALSE}
Map
```


## Outcome (performance) descriptive results

Next we can see if these categories of water systems perform differently using different performance measures related to water quality, accessibility and TMF capacity.


```{r describing performance outcomes by category, echo=FALSE, warning=FALSE, message=FALSE}

Outcomes <- Data[,c(8:14,27,28,30,35,38,39,41:43,47,48)]

Table2 <- Outcomes %>% tbl_summary(by = carceral_sys, missing = "no")

kable_Table2 <- Table2 %>% as_kable_extra(); kable_Table2
```

```{r describing performance outcomes by category 2, echo=FALSE, warning=FALSE, message=FALSE}
Outcomes_small <- Outcomes[,c(1,2,4,8,9,10,11,17)]

Table3 <- Outcomes_small %>% tbl_summary(by = carceral_sys, missing = "no")

kable_Table3 <- Table3 %>% as_kable_extra(); kable_Table3
```


## GLMs for outcomes with controls


Lastly we can use GLMs to see if any of these relationships are stastisticallly significant while holding other driving factors like system size constant. 


```{r GLMs, warning=FALSE, message=FALSE}
Data$carceral_sys <- relevel(Data$carceral_sys, ref = "other_water_systems")

MCL_violations <- glm(health_5yr_no2ndary_or_lsl_binary ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(MCL_violations)

M_violations <- glm(mon_only_5yr_binary ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(M_violations)

MR_violations <- glm(mr_5yr_binary ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(MR_violations)

NA_WQ <- lm(Water_Quality_Category_Score ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(NA_WQ)

NA_AC <- lm(Accessibility_Category_Score ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(NA_AC)

NA_TMF <- lm(TMF_Capacity_Category_Score ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(NA_TMF)

Data$Failingoratrisk <- ifelse(Data$SAFER_Status == "Failing" | Data$SAFER_Status == "At-Risk", 1, 0)
Data$failing <- ifelse(Data$SAFER_Status == "Failing", 1, 0)

Fail_or_atrisk <- lm(Failingoratrisk ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Fail_or_atrisk)

Fail <- lm(failing ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Fail)

CCR_anymissing <- glm(Missing_ccr_any ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(CCR_anymissing)

CCR_missingtwoplus <- glm(Missing_ccr_twoplus ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(CCR_missingtwoplus)

Modeleddroughtimpact <- glm(SC5e_Drought_Impact ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Modeleddroughtimpact)

Distribution_disrupt <- glm(Distributiondis_any ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Distribution_disrupt)

Sig_dif <- glm(Significant_Deficiencies ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Sig_dif)

Bottledwaterreliance <- glm(Bottled_Water_or_Hauled_Water_Reliance ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Bottledwaterreliance)

```

```{r GLMs 2, warning=FALSE, message=FALSE}
#Last thing, try EPA health and monitoring binary
EPA <- read_csv("Data/usepa_df_for_analysis.csv")
EPA <- EPA[,c(2,11,12)]
Data <- left_join(Data, EPA)

MCL_violations_epa <- glm(binary_viol_over5yrs_health ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(MCL_violations_epa)

MR_violations_epa <- glm(binary_viol_over5yrs_mon ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(MR_violations_epa)
```

```{r chisquared, warning=FALSE, message=FALSE}

chisq.test(Data$carceral_sys, Data$health_5yr_no2ndary_or_lsl_binary)
chisq.test(Data$carceral_sys, Data$mon_only_5yr_binary)
chisq.test(Data$carceral_sys, Data$Water_Quality_Category_Score)


CHIS <- lapply(Data[,-1], function(x) chisq.test(Data[,1], x)); CHIS
```

