---
title: "Final paper results Sept 24"
author: "Kristin Dobbin"
date: "2025-09-03"
output:
  html_document: default
knit: (function(rmdfile, ...) { rmarkdown::render(rmdfile, output_dir=here::here("Docs"))})
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = "/Users/KristinDobbin/Documents/Projects/R_Projects_Active/CAcarceralwatersystems", warning=FALSE, message=FALSE, include=FALSE)

library(tidyverse)
library(gtsummary)
library(kableExtra)
```


```{r get and prep data, include=FALSE, echo=FALSE}
Data <- read_csv("Data/Fullpaperdata_Sept24.csv")
Data <- Data %>% mutate(across(c(pwsid, pws_type_code, carceral_sys, health_5yr_no2ndary_or_lsl_binary, mon_only_5yr_binary, mr_5yr_binary, CASDWIS_County, watersource, purchased, SAFER_STATUS, SC5e_Drought_Impact, Distributiondis_any, HR_NAME, Water_Quality_Risk_Level_Display, Accessability_Risk_Level_Display, Affordability_Risk_Level_Display, TMF_Capacity_Risk_Level_Display), as.factor))
Data$Missing_ccr_any <- as.factor(ifelse(Data$n_ccr_missing > 0, "Yes", "No"))
Data$Missing_ccr_twoplus <- as.factor(ifelse(Data$n_ccr_missing >= 2, "Yes", "No"))
```

## Data overview

Our data set includes `r nrow(Data)` unique public water systems representing all of the active CWS and NTNCWS water systems in the same. Of these water systems `r sum(Data$pws_type_code == "CWS")` are community water systems, `r sum(Data$pws_type_code == "NTNCWS")` are Non-Transient Non-Community systems. The mean population served by these systems is `r mean(Data$population_served_count)`. 

## Describing carceral water systems

We classify these systems into three categories: Carceral water systems that primarily or exclusively serve carceral facilities;  non-exclusive carceral water systems that serve one or more carceral facilities in addition to other customers and non-carceral systems that do not serve any carceral faciltiies. We find that `r sum(Data$carceral_sys == "carceral_water_system")` or `r ((sum(Data$carceral_sys == "carceral_water_system"))/nrow(Data)*100)`% are carceral water systems.  `r sum(Data$carceral_sys == "carceral_serving_water_system")` or `r ((sum(Data$carceral_sys == "carceral_serving_water_system"))/nrow(Data)*100)`% are non-exclusive carceral systems and `r sum(Data$carceral_sys == "other_water_systems")` or `r ((sum(Data$carceral_sys == "other_water_systems"))/nrow(Data)*100)`% are noncarceral systems.

```{r describing carceral water system characteristics, echo=FALSE, warning=FALSE, message=FALSE}
System_descriptives <- Data[, c(3,6,8,17,18,19)]

Table <- System_descriptives %>% tbl_summary(by = carceral_sys)

kable_Table <- Table %>% as_kable_extra(); kable_Table
```

As you can see above, many of the carceral serving systems rely on surface water. Which is interesting. Many of those systems are concentrated in the bay area, Sacramento and northern central valley areas and southern california whereas exclusive carceral systems are fairly evenly spread throughout the state.

```{r describing carceral water system map, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(ggmap)
library(sf)
library(sp)

#load in boundary polygons for CWS
PWS_boundary <- st_read("Data/California_Drinking_Water_System_Area_Boundaries.geojson") #SABL download from Jenny 072325
str(PWS_boundary)
PWS_boundary <- st_make_valid(PWS_boundary)

PWS_boundary$SABL_PWSID <- as.factor(PWS_boundary$SABL_PWSID)
PWS_boundary$WATER_SYSTEM_NAME <- as.factor(PWS_boundary$WATER_SYSTEM_NAME)
PWS_boundary <- PWS_boundary %>% filter(BOUNDARY_TYPE != "Jurisdictional") #remove jurisdictional boundaries to get rid of duplicates

Geodata <- Data
Geodata <- left_join(Geodata, PWS_boundary, by = c("pwsid" = "SABL_PWSID"))
Geodata <- Geodata %>% distinct(pwsid, .keep_all = TRUE) #get rid of one duplicate

#get centroids
Geodata$Centroid <- NA
Geodata$Centroid <- st_centroid(Geodata$geometry) #Get centroids to service area boundaries

#Basemaps
caCountiesTmp <- tigris::counties(state = 06) %>%
  st_as_sf()

Geodata <- st_as_sf(Geodata, sf_column_name = "Centroid")

#maps - statewide
Map <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica"))

```

```{r system map, echo=FALSE, warning=FALSE, message=FALSE}
Map
```


## Outcome (performance) descriptive results

Next we can see if these categories of water systems perform differently using different performance measures related to water quality, accessibility and TMF capacity.


```{r describing performance outcomes by category, echo=FALSE, warning=FALSE, message=FALSE}

Outcomes <- Data[,c(8,9,11,23,24,26,28,37)]

Table2 <- Outcomes %>% tbl_summary(by = carceral_sys, missing = "no")

kable_Table2 <- Table2 %>% as_kable_extra(); kable_Table2
```

## Chi squared and GLMs for outcomes 

Lastly we can use GLMs and chi squared tests to see if any of these relationships are statistically significant while holding other driving factors like system size constant. 

```{r chisquared, warning=FALSE, message=FALSE}
MCL <- xtabs(~ health_5yr_no2ndary_or_lsl_binary + carceral_sys , data = Data)
chisq.test(MCL, correct = FALSE) #no sign dif between any groups

MON <- xtabs(~ mon_only_5yr_binary + carceral_sys , data = Data); MON
chisq.test(MON, correct = FALSE) #sig p=0.01163
chisq.test(MON[,c(2,1)]) #carceral serving and carceral sig different p=0.008
chisq.test(MON[,c(2,3)]) # carceral and other not sig (p=0.1342)
chisq.test(MON[,c(1,3)]) #carceral serving and other sig different p=0.01793

CCR <- xtabs(~ Missing_ccr_any + carceral_sys , data = Data); CCR
chisq.test(CCR, correct = FALSE) # Sig differences p<0.001
chisq.test(CCR[,c(2,1)]) # Carceral serving and carceral not different
chisq.test(CCR[,c(2,3)]) # Carceral and other systems not different
chisq.test(CCR[,c(1,3)]) #carceral serving and other systems sig dif p <0.001

Data_notassessedremoved <- Data
Data_notassessedremoved$Water_Quality_Risk_Level_Display[Data$Water_Quality_Risk_Level_Display == "Not Assessed"] <- NA
Data_notassessedremoved$Accessability_Risk_Level_Display[Data$Accessability_Risk_Level_Display == "Not Assessed"] <- NA
Data_notassessedremoved$TMF_Capacity_Risk_Level_Display[Data$TMF_Capacity_Risk_Level_Display == "Not Assessed"] <- NA
Data_notassessedremoved$SAFER_STATUS[Data$SAFER_STATUS == "Not Assessed"] <- NA
Outcomes_notassessedremoved <- Data_notassessedremoved[,c(8,9,11,23,24,26,28,37)]
Outcomes_notassessedremoved <- na.omit(Outcomes_notassessedremoved)
Outcomes_notassessedremoved <- droplevels(Outcomes_notassessedremoved)

QR <- xtabs(~ Water_Quality_Risk_Level_Display + carceral_sys , data = Outcomes_notassessedremoved); QR
chisq.test(QR, correct = FALSE) # not sig different p=0.4513
chisq.test(QR[,c(2,1)]) # not sig p=0.62
chisq.test(QR[,c(2,3)]) # not sig p=0.3091
chisq.test(QR[,c(1,3)]) #not sig p=0.7328

AR <- xtabs(~ Accessability_Risk_Level_Display + carceral_sys , data = Outcomes_notassessedremoved); AR
chisq.test(AR, correct = FALSE) # there are sig differneces, p<0.001
chisq.test(AR[,c(2,1)]) #Carceral serving and carceral sig different p<0.001
chisq.test(AR[,c(2,3)]) # carceral and other not sig different p=0.99
chisq.test(AR[,c(1,3)]) #carceral serving and other sig p<0.01

TMFR <- xtabs(~ TMF_Capacity_Risk_Level_Display + carceral_sys , data = Outcomes_notassessedremoved); TMFR
chisq.test(TMFR, correct = FALSE) # there are sig differneces, p<0.001
chisq.test(TMFR[,c(2,1)]) #Carceral serving and carceral sig different p<0.001
chisq.test(TMFR[,c(2,3)]) # carceral and other are sig different p<0.01
chisq.test(TMFR[,c(1,3)]) #carceral serving and other not sig p=0.1519

STATUS <- xtabs(~ SAFER_STATUS + carceral_sys , data = Outcomes_notassessedremoved); STATUS
chisq.test(STATUS, correct = FALSE) # there are sig differneces, p=0.00364
chisq.test(STATUS[,c(2,1)]) #Carceral serving and carceral sig different p=0.01139
chisq.test(STATUS[,c(2,3)]) # carceral and other are sig different p=004117
chisq.test(STATUS[,c(1,3)]) #carceral serving and other not sig different p=0.1166

```

GLMS have controls for population served, water source, if water is purchased and water system type as well as hydrologic region

```{r GLMs, warning=FALSE, message=FALSE}
Data$carceral_sys <- relevel(Data$carceral_sys, ref = "other_water_systems")

MCL_violations <- glm(health_5yr_no2ndary_or_lsl_binary ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(MCL_violations)

M_violations <- glm(mon_only_5yr_binary ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(M_violations)

MR_violations <- glm(mr_5yr_binary ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(MR_violations)

Data_notassessedremoved <- droplevels(Data_notassessedremoved)
Data_notassessedremoved_small <- Data_notassessedremoved[, c(8,23,24,26,28,6,17,18,3,19)]
Data_notassessedremoved_small <- Data_notassessedremoved_small %>% drop_na(Water_Quality_Risk_Level_Display, Accessability_Risk_Level_Display, TMF_Capacity_Risk_Level_Display, SAFER_STATUS)
Data_notassessedremoved_small$Water_Quality_Risk_Level_Display <- relevel(Data_notassessedremoved_small$Water_Quality_Risk_Level_Display, ref = "Low to no risk")
Data_notassessedremoved_small$Accessability_Risk_Level_Display <- relevel(Data_notassessedremoved_small$Accessability_Risk_Level_Display, ref = "Low to no risk")
Data_notassessedremoved_small$TMF_Capacity_Risk_Level_Display <- relevel(Data_notassessedremoved_small$TMF_Capacity_Risk_Level_Display, ref = "Low to no risk")
Data_notassessedremoved_small$carceral_sys <- relevel(Data_notassessedremoved_small$carceral_sys, ref = "other_water_systems")
Data_notassessedremoved_small$SAFER_STATUS <- factor(Data_notassessedremoved_small$SAFER_STATUS, ordered = TRUE, levels = c("Not At-Risk", "Potentially At-Risk", "At-Risk", "Failing"))

NA_WQ <- glm(Water_Quality_Risk_Level_Display ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data_notassessedremoved_small, family = binomial)
summary(NA_WQ)

NA_AC <- glm(Accessability_Risk_Level_Display ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data_notassessedremoved_small, family= binomial)
summary(NA_AC)

NA_TMF <- glm(TMF_Capacity_Risk_Level_Display ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data_notassessedremoved_small, family= binomial)
summary(NA_TMF)

library(MASS)
Data_notassessedremoved_small_nanone  <- na.omit(Data_notassessedremoved)
SAFER_STATUS <- polr(SAFER_STATUS ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data_notassessedremoved_small_nanone); SAFER_STATUS

CCR_anymissing <- glm(Missing_ccr_any ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(CCR_anymissing)

CCR_missingtwoplus <- glm(Missing_ccr_twoplus ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(CCR_missingtwoplus)

Modeleddroughtimpact <- glm(SC5e_Drought_Impact ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Modeleddroughtimpact)

Distribution_disrupt <- glm(Distributiondis_any ~ carceral_sys + population_served_count + watersource + purchased + pws_type_code + HR_NAME, data = Data, family= binomial)
summary(Distribution_disrupt)

```
Now I want to see if there are any geographical patterns based on performance - In order MCL violations, monitoring violations, quality risk, accessibility risk, TMF risk, safer status

```{r system map by outcomes, echo=FALSE, warning=FALSE, message=FALSE}
Geodata_MCL <- Geodata %>% filter(health_5yr_no2ndary_or_lsl_binary == 1)

Map_MCL <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata_MCL, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica")); Map_MCL

Geodata_MON <- Geodata %>% filter(mon_only_5yr_binary == 1)

Map_MON <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata_MCL, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica")); Map_MON

Geodata_QR <- Geodata %>% filter(Water_Quality_Risk_Level_Display == "High to medium risk")

Map_QR <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata_QR, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica")); Map_QR


Geodata_AR <- Geodata %>% filter(Accessability_Risk_Level_Display == "High to medium risk")

Map_AR <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata_AR, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica")); Map_AR

Geodata_TMFR <- Geodata %>% filter(TMF_Capacity_Risk_Level_Display == "High to medium risk")

Map_TMFR <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata_TMFR, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica")); Map_TMFR

Geodata_STATUS <- Geodata %>% filter(SAFER_STATUS == "Failing" | SAFER_STATUS == "At-Risk")

Map_STATUS <- ggplot() +
  geom_sf(data = caCountiesTmp, fill = "white") +
  geom_sf(mapping = aes(colour = carceral_sys, geometry = Centroid), size = .35, data = Geodata_STATUS, inherit.aes = FALSE) +
  scale_color_manual(values = c("carceral_water_system" = "goldenrod1", "other_water_systems" = "azure3", "carceral_serving_water_system" = "hotpink4")) +
  labs(color = "System Type") +
  theme(text = element_text(family = "Helvetica")); Map_STATUS

```

